---
title: "tidy-model"
author: "Jayson Ramos Vasquez"
date: "2025-02-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library('tidyverse') ; library('modeldata')
```

## Introduction

This R Markdown document demonstrates my abilities to use models for data analysis using a data set collected on crickets.

```{r load data, include = FALSE}
data(crickets, package = "modeldata")
names(crickets)
```

## What is that sound?

In this report we examine what factors predict a crickets chirp rate.

```{r summary, echo = FALSE}
summary(crickets)
ggplot(crickets, aes(x = rate)) + 
  geom_histogram(bins=8) + 
  ggtitle("Distribution of the chirp rate of crickets") + 
  xlab('Chirp rate (per min.)'  )
```

In this data set, there are two species of crickets with 31 observations split between the two species. The temperature ranges from 17.2 celsius to 30.4 celsisu. The mean rate of chirp rate is 72.89 chirps per minute.
## Temperature affects chirp rate

```{r temp, echo = FALSE}
ggplot(crickets, aes(x = temp, y = rate)) +
  geom_point() +
  geom_smooth(method = 'lm') +
  ggtitle("Plot of temperature and chirp rate") +
  ylab('Chirp rate(per min.)') +
  xlab('Temperature (Celsius)')

temp_lm <- lm(rate ~ temp, crickets)

summary.lm(temp_lm)
```

Based on a scatter plot of temperature and chirping and a correlation test, it seems that as temperature increases one degree, the rate of chirping increases about 4.2 chirps per minute. 

## Species-specific effects of temperature on chirping

```{r species, echo= FALSE}
ggplot(crickets, aes(x = temp, y = rate, color = species)) +
  geom_point() +
  geom_smooth(method = 'lm') +
  ggtitle("Plot of temperature and chirp rate for two species of crickets") +
  ylab('Chirp rate (per min.)') +
  xlab('Temperature (Celsius)')

species_lm <- lm(rate ~ temp + species, crickets)

summary.lm(species_lm)
```

This graph shows the difference between the two cricket species and their chirp rate based on temperature. With more data, it seems that as temperature increases one degree, the rate of chirping increases about 3.6 chirps per minute. Also, the species <i> O. exclamationis </i> has a chirp rate that is 10 chirps per minute higher than the species <i> O. niveus </i>. 

## Interactions

```{r species histogram, echo = FALSE}
ggplot(crickets, aes(x = rate, fill = species)) +
  geom_histogram(bins = 8) +
  ggtitle("Distribution of the chirp rate of crickets") +
  xlab('Chirp rate (per min.)')
```

This graph shows that these species occupy different temperature ranges.

```{r interactions, echo = FALSE}
ggplot(crickets, aes(x = temp, y = rate, color = species)) +
  geom_point() +
  geom_smooth(method = 'lm') +
  ggtitle("Plot of temperature and chirp rate for two species of crickets") +
  ylab('Chirp rate (per min.)') +
  xlab('Temperature (Celsius)')

species_x_temp_lm <- lm(rate ~ temp + species + temp:species, crickets)
summary.lm(species_x_temp_lm)

anova(species_lm, species_x_temp_lm)
```

I checked for interactions between species model and interaction model, but they are not significantly different, so I decided to stay with the species model. 